**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-09-03 submission/manuscript.Rmd
changese after R5 with Pat

__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS)[Figure 1].

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff

Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches

Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models and we highlighted potential pitfalls and ways to avoid them [Table S1].

The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1]. We benchmarked seven ML models and showed that if the goal is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome [Table S1]. We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance. We repeated this datasplit 100 times to get a robust reading of predictive performance. We tuned our model hyperparameters in a five-fold cross-validation. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. The feature weights of correlated OTUs are influenced by one another. For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight thus hiding the importance of its correlated OTU. To avoid misinterpreting the models and to capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we avoided misinterpreting models by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. To identify the true underlying microbial factors of a disease, it is crucial to do further experimentation for biological validation.

We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table S1] apply to all ML modeling tasks with microbiome data. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.

There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking. This would be an interesting direction for future work in microbiome studies.

Paragraph about deploying a model.

This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data. Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets. The right methods will help us achieve the level of validity and accountability we want from models built for patient health.

>>>>

__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].

@@

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations. Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process. In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification. However, the application of ML in microbiome studies is still in its infancy and the field still needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.

To address these needs, we developed a framework to train rigorous, transparent, and reproducible models. We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance. In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed a random datasplit to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this datasplit 100 times to measure predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. It is important to know the correlation strutures of the data to avoid missinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome. Therefore, to identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of categories to classify increase, more samples are required for each category to train a model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting. These methods are even more complex than random forest and XGBoost and are considered uninterpretable. There is great potential for applying ML approaches to microbiome data.

Our framework gives structure to investigators wanting to train, evaluate, and interpret their own ML models to identify OTUs that might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary we direct our efforts to overcome these challenges.

This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data. We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table 2]. We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation. Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.
